6.867 Problem Set 2

### 2015-10-27 HW2/tex/HW2.tex
[hw2] first draft of finished writeup.

This transformed model is known as logistic or logit regression. Using this formulation, our the negative log-likelihood becomes:
\begin{equation*}
	\textnormal{NLL}=\sum_i \log \left(1+e^{-y^{(i)}(w_0 + x^{(i)}\cdot w)} \right)
\end{equation*}
We think that overfitting might be problem in logistic regression especially in the case of linearly separable data which will drive our weights $w$ to very large values to minimize the the logistic loss. Hence, we add a ridge penalty to the weights on the features to shrink them:
\begin{equation*}
	\min_w \sum_i \log \left(1+e^{-y^{(i)}(w_0 + x^{(i)}\cdot w)} \right) + \lambda w^T w
\end{equation*}
Using this, we trained Logistic Regressions for each of our 4 datasets using gradient descent. The optimal weights are reported below:

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_stdev1_train.png}
		\caption*{stdev1 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_stdev2_train.png}
		\caption*{stdev2 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_stdev4_train.png}
		\caption*{stdev4 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_nonsep_train.png}
		\caption*{nonsep (Training)}
	\end{minipage}
		\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_stdev1_validation.png}
		\caption*{stdev1 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_stdev2_validation.png}
		\caption*{stdev2 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_stdev4_validation.png}
		\caption*{stdev4 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{LR_nonsep_validation.png}
		\caption*{nonsep (Validation)}
	\end{minipage}
	\caption{The decision boundaries generated by Logistic Regression plotted against the stdev1, stdev2, stdev4, and nonsep (training and validation) datasets}
\end{figure}

@@

We also explore the effect of adding a ridge penalty to logistic regression. We plot the classification error and loss of our regression as a function of the penalty $lambda$:

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{CErr_lambda_stdev1.png}
		\caption*{Error - stdev1}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{CErr_lambda_stdev2.png}
		\caption*{Error - stdev2}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{CErr_lambda_stdev4.png}
		\caption*{Error - stdev4}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{CErr_lambda_nonsep.png}
		\caption*{Error - nonsep}
	\end{minipage}
		\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Loss_lambda_stdev1.png}
		\caption*{Logistic Loss - stdev1}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Loss_lambda_stdev2.png}
		\caption*{Logistic Loss - stdev2}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Loss_lambda_stdev4.png}
		\caption*{Logistic Loss - stdev4}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{Loss_lambda_nonsep.png}
		\caption*{Logistic Loss - nonsep}
	\end{minipage}
	\caption{Classification Error (for a decision boundary of 0.5) and Loss of Logistic Regression as a function of the L2 penalty $\lambda$}
\end{figure}

@@

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_stdev1_train.png}
		\caption*{stdev1 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_stdev2_train.png}
		\caption*{stdev2 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_stdev4_train.png}
		\caption*{stdev4 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_nonsep_train.png}
		\caption*{nonsep (Training)}
	\end{minipage}
		\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_stdev1_validation.png}
		\caption*{stdev1 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_stdev2_validation.png}
		\caption*{stdev2 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_stdev4_validation.png}
		\caption*{stdev4 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\includegraphics[width=1\linewidth, height=1in]{linear_svm_nonsep_validation.png}
		\caption*{nonsep (Validation)}
	\end{minipage}
	\caption{The decision boundaries generated by SVM plotted against the stdev1, stdev2, stdev4, and nonsep (training and validation) datasets}
\end{figure}

@@

\subsubsection*{Titanic Data}
We're interesting in now testing the performance of our classifiers on a real dataset. In this case, we are using the Titanic dataset where we're doing the rather macabre task of trying to predict passenger survival given his or her features. So this problem we first scale all the features so they lie between $[0,1]$. WE accomplish by taking each observation and transforming it via:
\begin{equation*}
	x_{i,j}^* = \frac{x_{i,j}-\min(x_j)}{\max(x_j) - \min(x_j)}
\end{equation*}
We scale our data in such a way so our logistic regression behaves nicely. Otherwise, we encounter precision loss errors that end up significantly impacting our predictive accuracy.

Using this transformed data, we train a logistic regression, a linear SVM, a polynomial SVM ($K(x,z) = (1+x\cdot z)^3$), and a gaussian SVM. For each classifier, we perform a grid search to choose the optimal hyperparameters that minimize the classification error of the validation set. We find that for:
\begin{itemize}
	\item Logistic Regression, $\lambda = 0.38$
	\item Linear SVM, $C = 0.07$
	\item Polynomial SVM, $C=0.03$
	\item Gaussian SVM, $\sigma = 1$ and $C=0.04$
\end{itemize}

With these settings we obtain the following results:
\begin{table}[ht]
\centering
\captionof{table}{Titanic Data Classification Performance}
\begin{tabular}{lrrrr}
\toprule
{} & Logit & L-SVM & P-SVM & G-SVM \\
\midrule
Penalty & $\lambda = 0.38$ & $C = 0.07$ & $C = 0.03$ & $C = 0.04$\\
\midrule
Training Error    &  0.165 &  0.225 & 0.140 &  0.195 \\
Validation Error  &  0.208 &  0.208 & 0.222 &  0.210 \\
Test Error        &  0.249 &  0.233 & 0.265 &  0.243 \\
Geometric Margin  & {}     &  0.0055& 0.0112& 0.0063 \\
\bottomrule
\end{tabular}	
\end{table}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[height=3in]{titanic_weight_comparison.png}
	\caption*{Comparison of Feature Weights Across Models (Normalized per Model)}
\end{figure}

>>>>

This transformed model is known as logistic or logit regression. Using this formulation, our the negative log-likelihood becomes:
\begin{equation*}
	\textnormal{NLL}=\sum_i \log \left(1+e^{-y^{(i)}(w_0 + x^{(i)}\cdot w)} \right)
\end{equation*}
We think that overfitting might be problem in logistic regression especially in the case of linearly separable data which will drive our weights $w$ to very large values to minimize the the logistic loss. Hence, we add a ridge penalty to the weights:
\begin{equation*}
	\min_w \sum_i \log \left(1+e^{-y^{(i)}(w_0 + x^{(i)}\cdot w)} \right) + \lambda w^T w
\end{equation*}
Using this, we trained Logistic Regressions for each of our 4 datasets using gradient descent. The optimal weights are reported below:

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_stdev1_train.png}
		\caption*{stdev1 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_stdev2_train.png}
		\caption*{stdev2 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_stdev4_train.png}
		\caption*{stdev4 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_nonsep_train.png}
		\caption*{nonsep (Training)}
	\end{minipage}
		\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_stdev1_validation.png}
		\caption*{stdev1 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_stdev2_validation.png}
		\caption*{stdev2 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_stdev4_validation.png}
		\caption*{stdev4 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{LR_nonsep_validation.png}
		\caption*{nonsep (Validation)}
	\end{minipage}
	\caption{The decision boundaries generated by Logistic Regression plotted against the stdev1, stdev2, stdev4, and nonsep (training and validation) datasets}
\end{figure}

@@

We also explore the effect of adding a ridge penalty to logistic regression. We plot the classification error and loss of our regression as a function of the penalty $\lambda$:

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{CErr_lambda_stdev1.png}
		\caption*{Error - stdev1}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{CErr_lambda_stdev2.png}
		\caption*{Error - stdev2}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{CErr_lambda_stdev4.png}
		\caption*{Error - stdev4}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{CErr_lambda_nonsep.png}
		\caption*{Error - nonsep}
	\end{minipage}
		\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{Loss_lambda_stdev1.png}
		\caption*{Logistic Loss - stdev1}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{Loss_lambda_stdev2.png}
		\caption*{Logistic Loss - stdev2}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{Loss_lambda_stdev4.png}
		\caption*{Logistic Loss - stdev4}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{Loss_lambda_nonsep.png}
		\caption*{Logistic Loss - nonsep}
	\end{minipage}
	\caption{Classification Error (for a decision boundary of 0.5) and Loss of Logistic Regression as a function of the L2 penalty $\lambda$}
\end{figure}

@@

Table 3 shows the weights generated by our soft SVM algorithm, as well as its performance on the training and validation sets, given the four datasets we've been provided for testing. The soft SVM, in this case, uses a simple linear kernel, so the decision boundaries that are generated are still straight lines in $\mathbb{R}^2$. However, the slack provided by the soft SVM allows the algorithm to classify data and create a ``margin", even when the data is not linearly separable. As expected, in general the algorithm does slightly better on the training set than the validation set. Also, however, note that as the data becomes less and less linearly separable, our soft SVM with a linear kernel does increasingly poorly. This motivates our decision to explore the use of a Gaussian (or RBF) kernel.

Table 4 includes the performance of our soft SVM algorithm on various training sets when we change kernels (we provide data on both linear and Gaussian kernels), and also when we modify the bandwidth of our Gaussian kernel (where applicable) and change our regularization parameter, $C$. In general, our geometric margin increases as we decrease $C$. This is to be expected - as we provide the SVM with more slack, it is able to allow more points within its margin, and can consequently expand its geometric margin. In general, the larger the geometric margin, the more support vectors - again, this is intuitive. The more slack we allow, the more data points we will observe on or inside of our margin. It is not, however, necessarily true that the geometric margin will continue to increase as we decrease $C$ (and also, geometric margin will not always decrease as we increase $C$). As C gets very large, our decision boundary will approach the hard SVM decision boundary, and will not move at all. As $C$ gets increasingly small, the additional slack will lead our decision boundary to encompass all of the data, at which point our geometric margin will stop increasing.

Choosing $C$ by maximizing the geometric margin on the training data not a mistake that a practitioner should make. This is because maximizing the geometric margin on the training data is, in a sense, overfitting on the training data. Smaller $C$ would almost always give us larger geometric margins, and the practitioner would feel compelled to choose a very small $C$. However, this choice of $C$ may lead to very poor performance on validation sets, or future data. The best method for selecting $C$ would be to perform some sort of cross-validation. In the most basic sense, we might perform a grid search with various values of C, train our soft SVM, and see which value of C provides the best performance on a held-out validation set.

@@

\begin{figure}[ht]
	\centering
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_stdev1_train.png}
		\caption*{stdev1 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_stdev2_train.png}
		\caption*{stdev2 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_stdev4_train.png}
		\caption*{stdev4 (Training)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_nonsep_train.png}
		\caption*{nonsep (Training)}
	\end{minipage}
		\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_stdev1_validation.png}
		\caption*{stdev1 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_stdev2_validation.png}
		\caption*{stdev2 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_stdev4_validation.png}
		\caption*{stdev4 (Validation)}
	\end{minipage}
	\begin{minipage}[b]{.24\linewidth}
		\centering
		\includegraphics[width=.5\linewidth, height=.5in]{linear_svm_nonsep_validation.png}
		\caption*{nonsep (Validation)}
	\end{minipage}
	\caption{The decision boundaries generated by SVM plotted against the stdev1, stdev2, stdev4, and nonsep (training and validation) datasets}
\end{figure}

@@

\subsubsection*{Titanic Data}
We're interesting in now testing the performance of our classifiers on a real dataset. In this case, we are using the Titanic dataset, where we're performing the rather macabre task of predicting passenger survival given his or her features. For this problem we first scale all the features so they lie between $[0,1]$. We accomplish this by taking each observation and transforming it via:
\begin{equation*}
	x_{i,j}^* = \frac{x_{i,j}-\min(x_j)}{\max(x_j) - \min(x_j)}
\end{equation*}
This scaling of our data ensures that logistic regression behaves nicely. Otherwise, we encounter precision loss errors that end up significantly impacting our predictive accuracy.

Using this transformed data, we train a logistic regression, a linear SVM, a polynomial SVM ($K(x,z) = (1+x\cdot z)^3$), and a gaussian SVM. For each classifier, we perform a grid search to choose the optimal hyperparameters that minimize the classification error of the validation set. We find that for logistic regression,  $\lambda = 0.38$. For linear SVM, $C=0.07$. For polynomial SVM, $C=0.03$. And for Gaussian SVM, $\sigma = 1$ and $C=0.04$.

The results generated by these settings, as well as the weights generated by these settings, are found in Tables 5 and 6. Note that we omit our bias term, $w_0$, although it is included in the model, since it does not have any immediate physical interpretation (other than perhaps the baseline rate at which people survived the sinking of the Titanic).

As can be seen in Table 5, logistic regression outperforms polynomial SVM, with a test error rate of 24.9\% (as opposed to 26.5\% the SVM). Gaussian SVM performs outperforms the logistic regression, with a test error rate of 24.3\%, and the linear SVM performs the best, with a test error rate of 23.3\%. This is an interesting, and somewhat surprising result, as we might expect Gaussian SVM to perform as well (if not better) than linear SVM.

To better understand the differences in our models, one can look at Figure 5, which compares the weights of different variables for our different models. Note that each model's array of weights is normalized so that its maximum weight (in this case, always gender) is equal to 1. One can see that the linear SVM places almost all of its weight on gender, with a small amount on \# of parents/children. Logistic regression also places the most weight on gender, but also puts significant weight on (in order of decreasing magnitude) 3rd class passengers, age, \# of parents and children, 1st class passengers, and 2nd class passengers. The polynomial SVM puts most weight on passenger class and passenger embarkation locations. And the Gaussian SVM, after age, places the most weight on 3rd class passengers, 2nd class passengers, and passengers who embarked in Cherbourg.

\begin{table}[ht]
\centering
\captionof{table}{Titanic Data Classification Performance}
\begin{tabular}{lrrrr}
\toprule
{} & Logit & L-SVM & P-SVM & G-SVM \\
\midrule
Penalty & $\lambda = 0.38$ & $C = 0.07$ & $C = 0.03$ & $C = 0.04$\\
\midrule
Training Error    &  0.165 &  0.225 & 0.140 &  0.195 \\
Validation Error  &  0.208 &  0.208 & 0.222 &  0.210 \\
Test Error        &  0.249 &  0.233 & 0.265 &  0.243 \\
Geometric Margin  & {}     &  0.0055& 0.0112& 0.0063 \\
\bottomrule
\end{tabular}	
\end{table}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[height=3in]{titanic_weight_comparison.png}
	\caption*{Figure 5: Comparison of Feature Weights Across Models (Normalized per Model)}
\end{figure}
